### 2019-11-19 paper/conceptual_framework.tex
Edit and shorten conceptual framework

This section provides a conceptual framework for understanding what determines spacing 
between births.
Since sex selection is closely linked to education, I focus especially on the
the effects of education on spacing.
I first discuss a generic spacing decisions frameworks, and then turn to how son 
preference and sex selection fits into the decisions on spacing.

In the simplest possible dynamic model of fertility the spacing decision is trivial.
Assume that parents derive benefits from having their children around, rather than from 
the births themselves, that spacing of children by itself does impact parental utility, 
and that the parents perfectly control when to have children. 
In this set up, the optimal decision is then to have the children as early as possible 
\citep{Newman1988}.
The simplest model clearly does not fit the observed spacing patterns, but the question 
is why.

First, parents do not perfectly control the timing of births, but instead decide on their
levels of sexual activity and contraceptive effort, which, in turn, affect their likelihood 
of conceiving.

The decisions on sexual activity and contraceptive effort are moderated by fecundity,
determined by biological factors such as age and health of the mother, and the knowledge of and 
access to contraceptive technology.
There is evidence from ``natural fertility'' populations that increasing number of 
children is associated with shorter spacing and that birth intervals increase 
with parity \citep{Henry1961,Leridon1977}.

Biological factors are, however, unlikely to be an important factor in differences in spacing over 
time or across education levels in India for the period covered.
Outside of direct famine conditions, there is little evidence that women's health and 
nutritional status substantially affects spacing \citep{Huffman1987,John1987,lindstrom99}.%
\footnote{
On one hand, with better health we should expect shorter post-partum amenorrhoea and a higher
likelihood of carrying a pregnancy to term, both of which lead to shorter spacing with better health.
On the other hand, an healthier mother is more likely to successfully breastfeed and that would 
increase the spacing between births, although if she is a strong producer of breastmilk the 
suckling relative to production is lower, which, in turn, lower the conceptraceptive effect
of breastfeeding.
}
The Maharashtra drought occurred in 1970-1973 and, therefore, would only affect the first
year or two of the data.
Hence, although there are still differences in individual fecundity across women, it is
unlikely that we should see systematic differences across education groups and over time
in the data because of health status.
Furthermore, I only examine spacing up to parity fourth, and age of the mother is,
therefore, unlikely to be a concern when it comes to fecundity, when though better 
educated women are, on average, older when they marry.

Empirically, there is no clear relationship between contraception use and birth 
spacing or between education level and birth spacing
\citep{Tulasidhar1993,Whitworth2002,Bhalotra2008,Yeakey2009,Kim2010,Soest2018}.
The lack of evidence is consistent with the ambiguous theoretical effect of introducing 
modern contraceptives on birth spacing.
On one hand, when contraceptive technologies are less effective, parents may try to avoid 
overshooting their desired number of children by spacing their children further apart.
Increased reliability of access to and effectiveness of contraceptives can then lead to 
shorter intervals between births because parents are more certain of effectively
stopping childbearing when they reach their preferred number of children and having
the children closer together provide higher utility because parents get to enjoy the
children for longer as discussed above \citep{Keyfitz1971,Heckman1976}.
On the other hand, when contraceptives are less effective we would expect, through 
random variation, some birth intervals to be shorter than ideal for the reasons 
discussed below.
With increased reliability of access to and effectiveness of contraceptives, parents 
would better be able to avoid too short spacing between birth and average spacing
should go up.

To further complicate the relationship between contraception access and spacing,
the ability to successfully use ``low efficacy'' contraceptive methods, such as 
the rhythm method, is increasing in education, while there is no difference across education 
leves when it comes to modern contraceptives such as pills or injectables 
\citep{Rosenzweig1989}.
Furthermore, even in the absence of modern contraceptives, families appear able to time or 
at least postpone births when faced with strong enough incentives 
\citep{Jayachandran2011,Alam2018}.
[Does this need a tie-in to the paragraph above?]
[Can use "less effective" contraception and still have low fertility]
[Do parents move afway from certain contraceptive choices with education?]

Second, parents possibly care about the spacing between birth either in its own right or, 
more likely, because of the potential effects of spacing on the children and the mother.

It has, for example, been suggested that child ``quality'' might increases with spacing 
because the amount of parental attention increases with spacing
\citep{Zajonc1975,Zajonc1976,Razin1980}.
In this case, birth spacing should increase in family income, while the effect of maternal 
education is ambiguous since higher education on one hand makes the mother's time spent 
out of the labor market more costly and on the other hand increases how productive the 
mother is in childrearing. [TK Behrman et al. paper on mother's education in India]

One aspect of child ``quality'' where is is a strong effect of spacing is child health
and mortality, with most of the research finding that shorter spacing leads to worse
health and higher mortality for children, especially for very short intervals of 24 months 
or less \citep{Conde-Agudelo2006,Conde-Agudelo2012,Molitoris2019}.%
\footnote{
There is evidence in countries as diverse as Bangladesh, Brazil, El Salvadore, India,
and Pakistan of worse health outcomes and a higher mortality risk the shorter the birth
interval
\citep{Cleland1984,Curtis1993,Whitworth2002,Bhargava2003,Rutstein2005,Bhalotra2008,
Davanzo2008,Maitra2008,Makepeace2008,Gribble2009,Jayachandran2011,Saha2013,Jayachandran2017a,
Ghosh2018}.
This is different from the effect of the underlying mortality risk on birth spacing.
The expected effect of mortality risk---ignoring the effect of shorter spacing on mortality 
risk---is to shorten birth spacing to get to the ideal number of children earlier.
There is little empirical evidence for this type of effect 
\citep{Newman1983,Newman1988,Bhalotra2008}.
}
Hence, if parents realize that having children close together increase the children's
mortality risk, they would prefer to space births further apart to lower mortality risk
(although possibly not very long since the mortality risk involves a trade-off between the
cost of births and the benefit of having a child).
Furthermore, the effects of short spacing on child mortality are strongly moderated by
maternal education, and even very short spacing has relatively low mortality for women
with secondary or tertiary education \citep{Whitworth2002,Molitoris2019}.

The mother may also be affected by the length of the birth intervals, although the
results on mother's anthropometric status are mixed and there appear to be no
effect on maternal mortality \citep{Ronsmans1998,Menken2003,Dewey2007,Conde-Agudelo2012}.
For India, there is, however, evidence that women with first-born girls are more likely to 
have anemia, possibly because short birth spacing can resulting in maternal depletion 
\citep{Milazzo2018}.

Third, having and rearing children involves both direct and time costs, and economic factors 
can, therefore, affect birth spacing \citep{Hotz1997,Schultz1997}.
These economic factors can broadly be divided into 
the effects of price and income shocks,
the expected income trajectory of the husband and wife,
and economies of scale considerations.
There is a substantial amount of research on all these areas, although the majority is for
developed countries.

The expected effect of economies of scale in childrearing is to shorten the interval between 
births.
Childbearing and rearing is intensive in the mother's time, but taking care of two young
children is not twice as time consuming as taking care of one child 
\citep{Vijverberg1982,Espenshade1984}.
Hence, if childbearing and rearing requires the mother to reduce her participation in productive 
activities, parents can lower the cost of children by shorten birth spacing and, thereby, reduce
the amount of time the mother has to be away from productive activities.
In this set-up women with higher education will both have lower fertility and shorter
spacing than women with lower education because of the higher opportunity cost of time 
\citep{Ross1974,Newman1981,Newman1984}.
The extent to which this type of economies of scale is important depends on the extent to
which having children affects the mother's productive activities.
If she mainly works at home there likely is going to be less of an incentive to space 
children for this reason than if she is part of the labor force outside the home.
Similarly, higher household income, holding the mother's income contanst, reduces the incentive 
to space births for this reason.

In the absense of perfect capital markets, theory suggests that the steeper the parents' income 
profile over time, the longer the spacing between births 
\citep{Heckman1976,Wolpin1984,Newman1988}.
The idea is that although parents would prefer to have their children as early as possible to
enjoy them for longer there is also a direct cost to children.
At the early stage of life income is low, which means that the marginal utility of consumption 
is higher making parents less willing to give up consumption now to have a child, compare to 
later when they know that their income will be higher.
There is evidence that---at least for first births---there is, indeed, an effect of the income
profile over time \citep{Newman1984,Happel1984}.

There are two implications of this theoretical results.
First, we should see an general increase in spacing across education groups to the extent that
development lead to a steeper income profile.
[This does require that economic development is associated with steeper income profiles;
not sure if anybody looked at this.
The lower fertility even for low education women does suggest that income is higher,
but does not, by itself, say anything about the age profile in income.]
Second, the increase in spacing should be stronger with increasing education levels.
In other words, we should see a general increase in spacing, and this increase should be
larger the higher the education level.

@@

Finally, the discussion has so far assumed that children are identical, except for any potential
differences that arise from differences in spacing, but there is substantial evidence that the 
sex composition of previous children affect birth spacing.
For example, in Bangladesh, India, and Vietnam, having more boys, and particularly having 
a boy as the last-born child, significantly increase the duration to next birth,
\citep{Haughton1995,Haughton1996,Rahman1993,Bhalotra2008,Kumar2016,Soest2018}.%
\footnote{
There is also evidence of differential stopping behavior, with an additional child more likely the
fewer sons a family has 
\citep{repetto72,Das1987,Arnold1997,arnold98,clark00,Basu2010,Barcellos2014}.%
Other countries show a similar pattern
\citep[see, for example,][]{larsen98,filmer09,Altindag2016}.
Outside Asia, the evidence is more mixed, with North Africa showing shorter spacing in the 
absence of sons, while a similar pattern does not exist in Sub-Saharan Africa 
\citep{Rossi2015}.
}

It is not immediately obvious that we should expect a result like this from a theoretical
standpoint.
If, as discussed above, there are potentially significant negative effects of short spacing
parents who choose short spacing in the absence of a boy lowers the survival probabilty of
the next child, including a boy if the next child is a boy.
There are two possibilities for why parents might still try to conceive sooner in the 
absence of boys.%
\footnote{
The fact that there are significant differences in spacing by sex composition indicates that
parents have---at least some---control over the timing of birth.
Hence, it is unlikely that the differences in spacing come about by random chance.
In Matlab, the effect of son preference on birth spacing is stronger in areas with better 
access to family planning \citep{Rahman1993}.
}
First, it is possible that parents do not, in fact, realize that short spacing is potentially
detrimental to the next child's health.
Second, parents realize that short spacing has negative health effects, but have a strong
enough son preference that they are willing to divert substantial resources to investing
in the next child's health if it is a boy and that these investments can compensate for 
the negative effects.
[Are there any other ones that might explain the pattern?
    - biological? this would have to work off the sex of only the last child, so that
    spacing after girls is naturally shorter.
    - parents care so much about having a son sooner that they are willing to accept a 
    higher mortality risk.
    - ??
    ]

@@

[Need some kind of summary of spacing decisions and how they are likely to change
over time and with increased education]

As other countries like the US have developed, they have seen a progressively longer 
spacing between births as the proportion of births with very short intervals---less than 
24 months---have declined substantially \citep{Hotz1997}.
Most of the these countries do, however, not have the same degree of son preference that
we see in India.

What are the important factors that determine spacing?

- household income
- education
- labor force participation
- sex selection
- health (improvements, campaigns)

In summary, female education affects birth spacing through a variety of pathways.
The main pathway of interest here is the effect of sex selection on spacing.
However, to understand the effects of the increased use of sex selection over time it
is important to also understand the other changes that affect birth spacing.
The most likely factors are the changes in household income/wealth and the improvements
in health as India has developed.
I, therefore, now look at how female education and labor force participation has changed
over time in India.

>>>>

This section provides a conceptual framework for understanding what determines spacing 
between births.
I first discuss the main determinants of spacing discussed in the prior research, and 
then turn to how son preference and sex selection fits into the decisions on spacing.
Since sex selection is closely linked to education, I pay special attention to the
the effects of education on spacing.

In the simplest possible dynamic model of fertility the spacing decision is trivial.
Assume that parents derive benefits from the presence of children, rather than from 
the births themselves, that spacing of children by itself does impact parental utility, 
and that the parents perfectly control when to have children. 
In this set up, the optimal decision is then to have the children as early as possible 
\citep{Newman1988}.
The simplest model clearly does not fit the observed spacing patterns, which suggests 
that one or more of the assumptions in the simple model are wrong.

First, parents do not perfectly control the timing of births, but instead decide on the
levels of sexual activity and contraceptive effort, which, in turn, affect the 
likelihood of conceiving.

The decisions on sexual activity and contraceptive effort are moderated by fecundity,
determined by factors such as age and health of the mother, and the knowledge of and 
access to contraceptive technology.
There is evidence from ``natural fertility'' populations that increasing number of 
children is associated with shorter spacing and that birth intervals increase 
with parity \citep{Henry1961,Leridon1977}.

Age and health of the mother are, however, unlikely to be important factors in changes
in spacing over time or across education levels in India for the period covered, since
there is little evidence that women's health and nutritional status substantially affects 
spacing, outside of direct famine conditions \citep{Huffman1987,John1987,lindstrom99}.
The Maharashtra drought occurred in 1970--1973 and would, therefore, only affect the first
year or two of the data.
Furthermore, I only examine spacing up to parity fourth, and age of the mother is,
therefore, unlikely to be a concern when it comes to fecundity, even though better 
educated women are, on average, older when they marry.

Despite a general belief that better access to and knowledge of contraception should lead
to longer average spacing, the empirical relationship between contraception use and birth 
spacing is ambiguous \citep{Tulasidhar1993,Whitworth2002,Bhalotra2008,Yeakey2009,Kim2010,Soest2018}.
This lack of evidence is consistent with the ambiguous theoretical effect of introducing 
modern contraceptives on birth spacing.
On one hand, in the absence of modern contraceptive technologies, parents may try to 
avoid overshooting their desired number of children by spacing their children further apart.
Increased reliability of access to and effectiveness of contraceptives then lead to 
shorter intervals between births because parents are more certain of effectively
stopping childbearing when they reach their preferred number of children and having
the children closer together provide higher utility because parents benefit from the
presence of children for longer as discussed above \citep{Keyfitz1971,Heckman1976}.
On the other hand, when contraceptives are effective we would expect some birth 
intervals to be shorter than ideal in the sense discussed below through simple random
variation.
With increased reliability of access to and effectiveness of contraceptives, parents 
would better be able to avoid too short spacing between birth and average spacing
should go up.

To further complicate the relationship between contraception and spacing, the ability to 
successfully use ``low efficacy'' contraceptive methods, such as the rhythm method, is 
increasing in education, while there is no difference across education levels 
when it comes to modern contraceptives such as pills or injectables \citep{Rosenzweig1989}.
Furthermore, even in the absence of modern contraceptives, families appear able to time or 
at least postpone births when faced with strong enough incentives 
\citep{Jayachandran2011,Alam2018}.

Second, parents may care about the spacing between birth either in its own right or, 
more likely, because of the potential effects of spacing on the children.%
\footnote{
In principle, the mother could also be affected, but the results on mother's 
anthropometric status are mixed and there appear to be no effect on mortality 
\citep{Ronsmans1998,Menken2003,Dewey2007,Conde-Agudelo2012}.
}
It has, for example, been suggested that child ``quality'' might increases with spacing 
because the amount of parental attention increases with spacing
\citep{Zajonc1975,Zajonc1976,Razin1980}.
In this case, birth spacing should increase in family income, while the effect of maternal 
education is ambiguous since higher education on one hand makes the mother's time spent 
out of the labor market more costly and on the other hand increases how productive the 
mother is in childrearing. 
For child quality measures such as IQ and education there is, to my knowledge, no evidence 
from developing countries.%
\footnote{
The evidence from developed countries is mixed
\citep{Powell1993,Pettersson-Lidbom2009,Buckles2012,Barclay2017}.
}

For one crucial aspect of child quality---health and mortality---shorter spacing 
do lead to worse outcomes, especially for very short intervals of 24 months or less,
although this effect is strongly moderated by maternal education
\citep{Whitworth2002,Conde-Agudelo2006,Conde-Agudelo2012,Molitoris2019}.%
\footnote{
This is different from the potential effect of the underlying mortality risk on birth 
spacing, which---ignoring the effect of shorter spacing on mortality risk---is to shorten 
birth spacing to get to the ideal number of children earlier.
There is little empirical evidence for this effect \citep{Newman1983,Newman1988,Bhalotra2008}.
}
Hence, if parents realize short birth spacing increase their children's mortality risk, 
they likely prefer to space births further apart.
However, since women with secondary or tertiary education show relatively low mortality 
risk for their children, even with very short spacing is associated, this effect should
diminish with education level.

Third, having and rearing children involves both direct and time costs, and economic 
factors can, therefore, affect birth spacing \citep{Hotz1997,Schultz1997}.
The economic factors can broadly be divided into 
the effects of price and income shocks,
the expected income trajectory of the husband and wife,
and economies of scale considerations.

The expected effect of economies of scale in childrearing---childbearing and rearing is
time intensive, but taking care of two young children is not twice as time consuming as 
taking care of one child---is to shorten the interval between births
\citep{Vijverberg1982,Espenshade1984}.
To the extent that childbearing and rearing requires the mother to reduce her participation 
in productive activities, parents can lower the cost of children by shorten birth spacing 
thereby reducing the amount of time the mother is away from productive activities.
In this set-up, women with higher education will both have lower fertility and shorter
spacing than women with lower education because of the higher opportunity cost of time 
\citep{Ross1974,Newman1981,Newman1984}.
If the mother mainly works at home there is less incentive to space children for this 
reason and, similarly, higher household income, holding the mother's income contanst, 
reduces the incentive to space births for this reason.

In the absense of perfect capital markets, theory predicts that the steeper the parents'
life-time income profile, the longer the spacing between births, which suggests that 
with development we should see a general increase in spacing across education 
groups and that this increase in spacing should be stronger with increasing education levels
\citep{Heckman1976,Wolpin1984,Newman1988}.
The idea is that although parents would prefer to have their children as early as possible 
to enjoy them for longer, there is also a direct cost to children.
Early on income is low, which means that the marginal utility of consumption is higher 
making parents less willing to give up consumption now to have a child, compare to later 
when they know that their income will be higher \citep{Newman1984,Happel1984}.

@@

Finally, the discussion has so far assumed that children are identical---except for any 
potential differences that arise from differences in spacing---but there is substantial 
evidence that the sex composition of previous children affect birth spacing, with the
absence of boys significantly decreasing the duration to next birth
\citep{Haughton1995,Haughton1996,Rahman1993,Bhalotra2008,Kumar2016,Soest2018}.%
\footnote{
There is also evidence of differential stopping behavior, with an additional child more 
likely the fewer sons a family has 
\citep{repetto72,Das1987,Arnold1997,arnold98,clark00,Basu2010,Barcellos2014}.
}

It is not obvious that we should expect this result from a theoretical standpoint.
If, as discussed above, there are potentially significant negative effects of short spacing
parents who choose short spacing in the absence of a boy lowers the survival probabilty of
the next child, including a boy if the next child is a boy.
There are two possibilities for why parents might still try to conceive sooner in the 
absence of boys.%
\footnote{
The fact that there are significant differences in spacing by sex composition indicates that
parents have---at least some---control over the timing of birth.
Hence, it is unlikely that the differences in spacing come about by random chance.
In Matlab, the effect of son preference on birth spacing is stronger in areas with better 
access to family planning \citep{Rahman1993}.
}
First, it is possible that parents do not, in fact, realize that short spacing is potentially
detrimental to the next child's health.
Second, parents realize that short spacing has negative health effects, but have a strong
enough son preference that they are willing to divert substantial resources to investing
in the next child's health if it is a boy and that these investments can compensate for 
the negative effects.
[Are there any other ones that might explain the pattern?
    - biological? this would have to work off the sex of only the last child, so that
    spacing after girls is naturally shorter.
    - parents care so much about having a son sooner that they are willing to accept a 
    higher mortality risk.
    - ??
    ]

@@

In summary, female education affects birth spacing through a variety of pathways.
The main pathway of interest here is the effect of sex selection on spacing, but 
to fully understand how sex selection affects spacing we also need to understand 
the other changes that affect birth spacing over time.
The most likely factors are the changes in household income and wealth and the 
improvements in health as India has developed.
I, therefore, now look at how female education and labor force participation has changed
over time in India.
